---
title: Experimental evidence shows that bulb angle reflects changes in the angle of blow
substitle: Supplementary Information 1
author: "Li Li"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: no
    number_sections: yes
fontsize: 10pt
linestretch: 1.15
mainfont: Arial
---

```{r setup_analysis, include=FALSE}

library(ggpubr)
library(kableExtra)
library(ggplot2)
library(tidyverse)
library(reshape2)

knitr::opts_chunk$set(
	echo = FALSE,
	fig.pos = "!H",
	message = FALSE,
	warning = FALSE,
	cache = FALSE,
	# comment this to show hashtag for code output
	comment = NA,
	prompt = FALSE
)

# colorblind palette
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")


# the drop tower dataset
drop_tower_data = read.csv("ESM_2.csv")

# the Dibble dataset with bulb angle measured
dibble_data = read.csv("ESM_3.csv")

# the MPI dataset
mpi_data = read.csv("ESM_4.csv")

# flakes used in the EPA-PD model
# same core morphology and no beveling
dibble_lm = dibble_data %>%
  filter(coretype == "SEMISPHERICAL") %>%
  filter(bevel == "NONE") %>%
  mutate(fweight_cbrt = as.numeric(fweight)^(1/3)) %>%
  filter(fweight > 0)

new_dibble_bulb = melt(dibble_data, id.vars = "aob") %>%
  filter(variable == "bulb_angle_vc" | variable == "bulb_angle_vg")

```

```{r}

## function used to save plots
save_plot = function(plot_name, width, height){
  code_path = getwd()
  dir.create(file.path(code_path, 'figures'), showWarnings = FALSE)
  plot_path = paste(code_path, "/figures/", deparse(substitute(plot_name)), ".pdf", sep = "")
  return(ggsave(plot_path, plot = plot_name, width = width, height = height))
}

```

# The drop tower setup
The drop tower used in our experiment is made of a vertical stand with an adjustable pole. An electromagnet switch controls the release of the hammer attached to a lever that can move along the supporting pole to adjust the hammer’s drop height. To test the precision of the hammer strike, we dropped ball bearings onto targets drawn on carbon paper and compared the position of the mark made by the ball bearing and the original targets. The drop tower setup can achieve a precision of around 2 mm for aiming the striking target.

# Figures

```{r, fig.align='center', fig.width = 6, fig.height = 4}

fig6 = ggplot(drop_tower_data, aes(x = factor(AOB), y = bulb_angle)) +
  geom_boxplot() +
  geom_jitter(position = position_dodge2(width = 0.3)) +
  theme_classic() +
  labs(x = "AOB (degrees)",
       y = "Bulb angle (degrees)")

fig6

save_plot(fig6, 6, 4)

```

__Fig.6 Boxplot showing that bulb angle decreases as angle of blow increases for flakes in the drop tower dataset.__

```{r, fig.align='center', fig.width = 6, fig.height = 4}

fig7 = ggplot(new_dibble_bulb, 
                   aes(x = factor(aob), y = as.numeric(value), fill = variable)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(fill = "Method", 
       x = "AOB (degrees)",
       y = "Bulb angle (degrees)") +
  scale_fill_manual(values = cbPalette, 
                    labels = c("Vector calculation", "Virtual goniometer")) +
  ylim(125, 165)

fig7

save_plot(fig7, 6, 4)

```

__Fig.7 Boxplot summarizing the relationship between bulb angle and angle of blow for flakes in the Dibble dataset. Flakes measured with the vector calculation method are colored in grey, flakes measured with the virtual goniometer method are colored in yellow.__


```{r fig.align='center', fig.width = 5, fig.height = 3.5}

fig8 = ggplot(dibble_data, aes(x = bulb_angle_vc, y = bulb_angle_vg, color = factor(aob))) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  theme_classic() +
  scale_color_viridis_d(name = "AOB") +
  xlim(125, 165) +
  ylim(125, 165) +
  xlab("Bulb angle by vector calculation (degrees)") +
  ylab("Bulb angle by virtual goniometer (degrees)")

fig8

save_plot(fig8, 5, 4)

```

__Fig.8 Comparison of the flake bulb angles measured with the vector calculation method and the virtual goniometer method in the Dibble dataset.__  


```{r, fig.align='center', fig.width = 6, fig.height = 4}

fig9 = ggplot(data = mpi_data, aes(Bulb_angle)) +
  geom_histogram(aes(y = ..density..),
    binwidth = 1, colour="black", fill="grey") +
  geom_density(alpha = .2, fill = "antiquewhite3") +
  theme_classic() +
  labs(x = "Bulb angle (degrees)",
       y = "Density")

fig9

save_plot(fig9, 6, 4)

```
__Fig.9 Histogram showing the distribution of bulb angles with overlaying density curve for all flakes analyzed in the MPI dataset.__  


```{r echo=FALSE, fig.width = 6, fig.height = 4, mpi_all}

mean_mpi_ba = mean(mpi_data$Bulb_angle)

mpi_data = mpi_data %>%
  mutate(AOB_predicted = ifelse(Bulb_angle <= mean_mpi_ba, "High", "Low"))

fig10 = ggplot(mpi_data, aes(x = AOB_assigned, y = Bulb_angle)) +
  geom_boxplot(outlier.shape=NA) +
  geom_jitter(position = position_dodge2(width = 0.2)) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(x = "AOB assigned by knappers",
       y = "Bulb angle (degrees)")

fig10

save_plot(fig10, 6, 4)

```

__Fig.10 Boxplot showing the result of the angle of blow blind test with the MPI dataset.__  


```{r}

# the baseline EPA-PD model
dibble_lm_epa_pd = lm(fweight_cbrt ~ platthick + epa, data = dibble_lm)
#summary(dibble_lm_epa_pd)

# the EPA-PD-AOB model
dibble_lm_epa_pd_aob = lm(fweight_cbrt ~ platthick + epa + aob, data = dibble_lm)
#summary(dibble_lm_epa_pd_aob)

# the EPA-PD-BA model
dibble_lm_epa_pd_vc = lm(fweight_cbrt ~ platthick + epa + bulb_angle_vc, data = dibble_lm)
#summary(dibble_lm_epa_pd_vc)

dibble_lm_epa_pd_vg = lm(fweight_cbrt ~ platthick + epa + bulb_angle_vg, data = dibble_lm)
#summary(dibble_lm_epa_pd_vg)

## add modeled mass for plotting
dibble_lm$mass_pred1 = predict(dibble_lm_epa_pd, data = dibble_lm)
dibble_lm$mass_pred2 = predict(dibble_lm_epa_pd_aob, data = dibble_lm)
dibble_lm$mass_pred3 = predict(dibble_lm_epa_pd_vc, data = dibble_lm)
dibble_lm$mass_pred4 = predict(dibble_lm_epa_pd_vg, data = dibble_lm)

```

```{r fig.align='center', fig.width = 10, fig.height = 11}

fig11a = ggplot(dibble_lm, aes(x = fweight_cbrt, y = mass_pred1)) +
  geom_point(aes(color = factor(aob))) +
  geom_smooth(formula = y ~ x, method = "lm") +
  theme_classic() +
  scale_color_viridis_d() +
  labs(color = "AOB") +
  xlab("Actual flake mass (cube root)\nBasic EPA-PD model") +
  ylab("Predicted flake mass (cube root)")

fig11b = ggplot(dibble_lm, aes(x = fweight_cbrt, y = mass_pred2)) +
  geom_point(aes(color = factor(aob))) +
  geom_smooth(formula = y ~ x, method = "lm") +
  theme_classic() +
  scale_color_viridis_d() +
  labs(color = "AOB") +
  xlab("Actual flake mass (cube root)\nEPA-PD-AOB model") +
  ylab("Predicted flake mass (cube root)")

fig11c = ggplot(dibble_lm, aes(x = fweight_cbrt, y = mass_pred3)) +
  geom_point(aes(color = factor(aob))) +
  geom_smooth(formula = y ~ x, method = "lm") +
  theme_classic() +
  scale_color_viridis_d() +
  labs(color = "AOB") +
  xlab("Actual flake mass (cube root)\nEPA-PD-BA (VC) model") +
  ylab("Predicted flake mass (cube root)")

fig11d = ggplot(dibble_lm, aes(x = fweight_cbrt, y = mass_pred4)) +
  geom_point(aes(color = factor(aob))) +
  geom_smooth(formula = y ~ x, method = "lm") +
  theme_classic() +
  scale_color_viridis_d() +
  labs(color = "AOB") +
  xlab("Actual flake mass (cube root)\nEPA-PD-BA (VG) model") +
  ylab("Predicted flake mass (cube root)")

fig11 = ggarrange(fig11a, fig11b, fig11c, fig11d, nrow = 2, ncol = 2, labels = c("(a)", "(b)", "(c)", "(d)"), vjust = 1, hjust = 0, common.legend = TRUE)

fig11
save_plot(fig11, 7, 8)

```

__Fig.11 Comparison of the actual flake mass and modeled flake mass in the Dibble dataset, n = 53. (a) Actual to predicted flake mass using the basic EPA-PD model, (b) actual to predicted flake mass using the EPA-PD-AOB model, (c) actual to predicted flake mass using the EPA-PD-BA (vector calculation) model, and (d) actual to predicted flake mass using the EPA-PD-BA (virtual goniometer) model.__

# Tables
__Table 1 Summary statistics of the different EPA-PD models__  

```{r}
#summary(dibble_lm_epa_pd)
dibble_lm_r_sqrd = round(summary(dibble_lm_epa_pd)$r.squared,2)
dibble_lm_r_sqrdadj = round(summary(dibble_lm_epa_pd)$adj.r.squared,2)
dibble_lm_f_value = round(summary(dibble_lm_epa_pd)$fstatistic[1],2)
dibble_lm_df = summary(dibble_lm_epa_pd)$fstatistic[2]
dibble_lm_dendf = summary(dibble_lm_epa_pd)$fstatistic[3]

#summary(dibble_lm_epa_pd_aob)
dibble_lm_aob_r_sqrd = round(summary(dibble_lm_epa_pd_aob)$r.squared,2)
dibble_lm_aob_r_sqrdadj = round(summary(dibble_lm_epa_pd_aob)$adj.r.squared,2)
dibble_lm_aob_f_value = round(summary(dibble_lm_epa_pd_aob)$fstatistic[1],2)
dibble_lm_aob_df = summary(dibble_lm_epa_pd_aob)$fstatistic[2]
dibble_lm_aob_dendf = summary(dibble_lm_epa_pd_aob)$fstatistic[3]

## vector calculation method
#summary(dibble_lm_epa_pd_vc)
dibble_lm_vc_r_sqrd = round(summary(dibble_lm_epa_pd_vc)$r.squared,2)
dibble_lm_vc_r_sqrdadj = round(summary(dibble_lm_epa_pd_vc)$adj.r.squared,2)
dibble_lm_vc_f_value = round(summary(dibble_lm_epa_pd_vc)$fstatistic[1],2)
dibble_lm_vc_df = summary(dibble_lm_epa_pd_vc)$fstatistic[2]
dibble_lm_vc_dendf = summary(dibble_lm_epa_pd_vc)$fstatistic[3]

## virtual goniometer method
#summary(dibble_lm_epa_pd_vg)
dibble_lm_vg_r_sqrd = round(summary(dibble_lm_epa_pd_vg)$r.squared,2)
dibble_lm_vg_r_sqrdadj = round(summary(dibble_lm_epa_pd_vg)$adj.r.squared,2)
dibble_lm_vg_f_value = round(summary(dibble_lm_epa_pd_vg)$fstatistic[1],2)
dibble_lm_vg_df = summary(dibble_lm_epa_pd_vg)$fstatistic[2]
dibble_lm_vg_dendf = summary(dibble_lm_epa_pd_vg)$fstatistic[3]

dibble_lm_stats = data.frame(
  basic_lm = c(dibble_lm_r_sqrd, dibble_lm_r_sqrdadj, paste(dibble_lm_f_value,"(",dibble_lm_df,",",dibble_lm_dendf,")", sep = "")),
  aob_lm = c(dibble_lm_aob_r_sqrd, dibble_lm_aob_r_sqrdadj, paste(dibble_lm_aob_f_value,"(",dibble_lm_aob_df,",",dibble_lm_aob_dendf,")", sep = "")),
  vc_lm = c(dibble_lm_vc_r_sqrd, dibble_lm_vc_r_sqrdadj, paste(dibble_lm_vc_f_value,"(",dibble_lm_vc_df,",",dibble_lm_vc_dendf,")", sep = "")),
  vg_lm = c(dibble_lm_vg_r_sqrd, dibble_lm_vg_r_sqrdadj, paste(dibble_lm_vg_f_value,"(",dibble_lm_vg_df,",",dibble_lm_vg_dendf,")", sep = ""))
)

rownames(dibble_lm_stats) = c("R$^2$","Adj. R$^2$","F statistics")

kbl(
  dibble_lm_stats,
  booktabs = T, 
  align = "c",
  escape = F,
  row.names = T,
  col.names = linebreak(c("Basic EPA-PD model", "EPA-PD-AOB model", "EPA-PD-BA (VC) model",  "EPA-PD-BA (VG) model"), align = "l")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position")) %>%
  column_spec(1, width = "1.8cm") %>%
  column_spec(2:5, width = "3cm")

```

# Supplementary Figures

```{r, fig.align='center', fig.width = 6, fig.height = 4}

# calculate the standard error of bulb angle measurements in the drop tower dataset
drop_error = drop_tower_data %>%
  rowwise() %>%
  mutate(ba_sd = round(sd(c(bulb_angle_1, bulb_angle_2, bulb_angle_3)), 2),
         ba_se = round(ba_sd/sqrt(4), 2))

mean_se_1 = mean(drop_error$ba_se)

fig4.1 = ggplot(data = drop_error, aes(ba_se)) +
  geom_histogram(binwidth = 0.2, colour="black", fill="grey") +
  theme_classic() +
  labs(x = "Standard error of bulb angle (degrees)",
       y = "Count")

fig4.1

#save_plot(fig4.1, 6, 4)
```
__Fig4.1 Histogram showing the distribution of the standard error of the three bulb angle measurements for each flake in the drop tower dataset.__  

```{r, fig.align='center', fig.width = 6, fig.height = 4}

# calculate the standard error of bulb angle measurements in the drop tower dataset
mpi_error = mpi_data %>%
  rowwise() %>%
  mutate(ba_sd = round(sd(c(bulb_angle_1, bulb_angle_2, bulb_angle_3, bulb_angle_4)), 2),
         ba_se = round(ba_sd/sqrt(4), 2))

mean_se_2 = mean(mpi_error$ba_se)

fig4.2 = ggplot(data = mpi_error, aes(ba_se)) +
  geom_histogram(binwidth = 0.2, colour="black", fill="grey") +
  theme_classic() +
  labs(x = "Standard error of bulb angle (degrees)",
       y = "Count")

fig2.2

#save_plot(fig4.2, 6, 4)
```
__Fig4.1 Histogram showing the distribution of the standard error of the four bulb angle measurements for each flake in the MPI dataset.__  

\pagebreak

# Supplementary Tables
__Table 5.1 Summary statistics of bulb angle by angle of blow for the drop tower dataset (n = 103).__

```{r}

drop_tower_data %>%
  group_by(AOB) %>%
  summarise(count = n(),
            mean = round(mean(bulb_angle,na.rm=TRUE), 1), 
            sd = round(sd(bulb_angle, na.rm = TRUE), 2),
            var = round(var(bulb_angle, na.rm = TRUE), 2)) %>%
            #se = round(sd/sqrt(count), 2)) %>%
  kbl(align = "c",
      col.names = linebreak(c("AOB", "Count", "Bulb angle (mean)", "SD", "Variance"), align = "l"),
      booktabs = T) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                latex_options = "HOLD_position")

```

__Table 5.2 Summary of bulb angle by angle of blow based on the measurement method, VG refers to the virtual goniometer method and VC refers to the vector calculation method (the Dibble dataset, n = 70).__  

```{r tab-2}

dibble_data %>%
  group_by(aob) %>%
  summarise(mean_vector = round(mean(bulb_angle_vc,na.rm=TRUE), 1), 
            sd_vector = round(sd(bulb_angle_vc, na.rm = TRUE), 1),
            #var_vector = round(var(bulb_angle_vc, na.rm = TRUE), 1),
            mean_v_goniometer = round(mean(bulb_angle_vg,na.rm=TRUE), 1), 
            sd_v_goniometer = round(sd(bulb_angle_vg, na.rm = TRUE), 1),
            #var_v_goniometer = round(var(bulb_angle_vg, na.rm = TRUE), 1),
            count = n()
            ) %>%
  kbl(col.names = linebreak(c("AOB", "Bulb angle (mean)", "SD", "Bulb angle (mean)", "SD", "Count"), align = "l"),
      booktabs = T) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                latex_options = "HOLD_position") %>%
   add_header_above(c(" " = 1, "Vector calculation" = 2, "Virtual goniometer" = 2, " " = 1))

```

__Table 5.3 Summary of angle blow prediction for all flakes in the MPI dataset.__    

```{r}

mpi_data %>%
  group_by(AOB_assigned) %>%
  summarise(mean_aob = round(mean(Bulb_angle), 1),
            sd_ba = round(sd(Bulb_angle), 2),
            count = n()) %>%
  kbl(col.names = linebreak(c("AOB assigned by knappers", "Average bulb angle", "SD of bulb angle", "Count"), align = "l"),
      booktabs = T) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE,
                latex_options = "HOLD_position")

```

\pagebreak

# Statistical analysis

## The drop tower dataset
__Kruskal-Wallis test checking the difference in bulb angle by angle of blow for flakes in the drop tower dataset:__  
```{r}

## Kruskal-Wallis test of the drop tower dataset
kruskal.test(bulb_angle ~ factor(AOB), data = drop_tower_data)

```

__Linear regression to examine the effect of angle of blow and platform depth on bulb angle for flakes in the drop tower dataset:__  
```{r}

summary(lm(bulb_angle ~ AOB + pd_mm, data = drop_tower_data))

```

## The Dibble dataset

__Kruskal-Wallis test checking the difference in bulb angle (measured with the vector calculation method) by angle of blow for flakes in the Dibble dataset:__  
```{r}

# Krukal-Wallis test on flakes measured with the vector calculation method
kruskal.test(bulb_angle_vc ~ factor(aob), data = dibble_data)

```

__Linear regression to examine the effect of angle of blow, exterior platform angle, and platform depth on bulb angle (measured with the vector calculation method) for flakes in the Dibble dataset:__  
```{r}

summary(lm(bulb_angle_vc ~ aob + platthick + epa, data = dibble_data))

```

__Linear regression to examine the correlation between the angle of blow and platform depth in the Dibble dataset:__  

```{r}

# there is a significant correlation between platform depth and angle of blow
summary(lm(platthick ~ aob, data = dibble_data))

```

__Kruskal-Wallis test checking the difference in bulb angle (measured with the virtual goniometer method) by angle of blow for flakes in the Dibble dataset:__  
```{r}

# Krukal-Wallist test on flakes measured with the virtual goniometer method
kruskal.test(bulb_angle_vg ~ factor(aob), data = dibble_data)

```

__Linear regression to examine the effect of angle of blow, exterior platform angle, and platform depth on bulb angle (measured with the virtual goniometer method) for flakes in the Dibble dataset:__  
```{r}

summary(lm(bulb_angle_vg ~ aob + epa + platthick, data = dibble_data))

```


## The MPI dataset

__Kruskal-Wallis tests comparing the difference in bulb angle by angle of blow (the MPI dataset):__  
```{r}

## Kruskal-Wallis test of the difference in bulb angle by angle of blow predicted by LL for the MPI dataset:
kruskal.test(Bulb_angle ~ factor(AOB_assigned), data = mpi_data)

```

## The EPA-PD model comparison

__ANOVA test comparing the EPA-PD model and the EPA-PD-AOB model:__  

```{r}

anova(dibble_lm_epa_pd, dibble_lm_epa_pd_aob)

```

__ANOVA test comparing the EPA-PD model and the EPA-PD-BA model:__  

```{r}

anova(dibble_lm_epa_pd, dibble_lm_epa_pd_vc)
anova(dibble_lm_epa_pd, dibble_lm_epa_pd_vg)

```
